Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-09 treestats_paper.tex
what is a branch stat from real data?

@@

The comparisons are shown in Figure~\ref{fig:chr20}.
Of these methods, branch diversity from tsinfer shows the worst agreement with site diversity,
which is unsurprising as tsinfer makes no attempt to provide well-calibrated node times,
and only aims to provide consistent tree topologies.
The two methods that infer node times differ substantially,
with Relate showing a significantly tighter agreement than GEVA.
Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} plus GEVA?
The results presented here are highly suggestive, but not entirely clear.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.region.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} tsinfer,
        \textbf{(bottom left)} Relate, and
        \textbf{(bottom right)} GEVA.
        \label{fig:chr20}
    }
\end{figure}

\krt{Hmmm...not sure about this paragraph.  Under strict BGS, some of what is stated here doesn't hold.  The tree will
    still be neutral in shape, just shorter overall.  For the case of many weakly-deleterious segregating mutations, the
    overall shape of the tree may not be that different from neutral expectations (Neuhaueser/Krone and the later
Przeworksi paper, all in Genetics IIRC?) It also seems to me that the idea of what branch-vs-site should look like under
these processes vs what they should look like in a ts inferred from data where the true model is these processes are
getting conflated.}
In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{background_selection}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does not affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
However, it is generally unknown how much mutation rate varies along the genome
\citep[but see][]{mut_rate_varies}
or how dense are targets of selection \citep{how_important_is_selection}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by selection and/or mutation rate variation.

>>>>

\krt{Is there a circularity here?  What does a branch statistic from real data mean
when the branch lengths must be inferred somehow from the mutations? If this is not circular,the 
reasons why need to be made clearer.}

@@

The comparisons are shown in Figure~\ref{fig:chr20}.
Of these methods, branch diversity from tsinfer shows the worst agreement with site diversity,
which is unsurprising as tsinfer makes no attempt to provide well-calibrated node times,
and only aims to provide consistent tree topologies.
The two methods that infer node times differ substantially,
with Relate showing a significantly tighter agreement than GEVA.
Does this mean that Relate is doing a better job at inferring the tree sequence than \texttt{tsinfer} pluse GEVA?
The results presented here are highly suggestive, but not entirely clear.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.region.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        \krt{the only appearances of "dual branch" are in figure legends.  need to define somewhere}
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} tsinfer,
        \textbf{(bottom left)} Relate, and
        \textbf{(bottom right)} GEVA.
        \label{fig:chr20}
    }
\end{figure}

\krt{Not sure exactly how to rewrite.}
In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{background_selection}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does not affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
However, it is generally unknown how much mutation rate varies along the genome
\citep[but see][]{mut_rate_varies}
or how dense are targets of selection \citep{how_important_is_selection}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by selection and/or mutation rate variation.
